---
title: "Portfolio 5: Interactive Spotify Track Explorer"
---

>Goal 
I wanted to try my luck and create a small interactive data tool using R Shiny. I am using a Spotify Tracks dataset from Kaggle. Instead of making one static graph, I wanted users to be able to choose a music genre and compare different Spotify audio features. My goal is to practice building something that feels more like a usable product or mini-dashboard, while still keeping the project simple enough to finish for my portfolio.

Here is the link to the Kaggle dataset page: https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset?select=dataset.csv

Each row in the dataset represents one track. The dataset includes song information such as track name, artist, album, genre, and popularity. It also includes audio features such as danceability, energy, loudness, speechiness, acousticness, instrumentalness, liveness, valence, and tempo.

The main question I explored was:

**How do Spotify track features, such as danceability, energy, valence, tempo, and popularity, differ across genres?**

## Product

The final product is an interactive R Shiny explorer. A user can:

1. Choose a genre.
2. Choose an x-axis feature.
3. Choose a y-axis feature.
4. Filter tracks by popularity.
5. View a plot that updates automatically.
6. See the top songs in the selected genre.

### Load packages 

```{r seup}
library(tidyverse)
library(shiny)
library(DT)
```

### Import dataset

```{r}
spotify_raw <- read_csv("data/dataset.csv", show_col_types = FALSE)

glimpse(spotify_raw)
```

### Clean the data

I am going to start by selecting the columns that are most useful for the app.
I want to keep the app simple and avoids overwhelming the user with too many variables.

```{r clean data}
spotify <- spotify_raw %>%
  select(
    track_name,
    artists,
    album_name,
    popularity,
    duration_ms,
    explicit,
    danceability,
    energy,
    loudness,
    speechiness,
    acousticness,
    instrumentalness,
    liveness,
    valence,
    tempo,
    track_genre
  ) %>%
  drop_na(track_name, artists, popularity, danceability, energy, valence, tempo, track_genre)
```

I will also convert the duration from milliseconds to minutes because minutes are easier to interpret.

```{r convert duration}
spotify <- spotify %>%
  mutate(
    duration_min = duration_ms / 60000,
    explicit = if_else(explicit == TRUE, "Explicit", "Not explicit")
  )
```

### Create a list of features users can choose from

These will be the numeric features users will be able to compare in the app from the menu. I am choosing features that are meaningful for music exploration.

The names on the left are what the user sees. The names on the right are the actual column names in the dataset.

```{r feature-list}
feature_choices <- c(
  "Popularity" = "popularity",
  "Danceability" = "danceability",
  "Energy" = "energy",
  "Valence / Positivity" = "valence",
  "Tempo" = "tempo",
  "Acousticness" = "acousticness",
  "Instrumentalness" = "instrumentalness",
  "Speechiness" = "speechiness",
  "Liveness" = "liveness",
  "Loudness" = "loudness",
  "Duration in minutes" = "duration_min"
)
```

I am hoping for the final product to be an interactive Spotify track explorer. The user can select a music genre, choose two audio features to compare, and filter songs by popularity. The app updates the scatterplot and table based on the user's choices.

This project helped me practice thinking about data from a user perspective. Instead of only creating one static visualization, I created a small tool that allows users to explore the dataset themselves.

### Build the Shiny app interface

This took some work :/ (hopefully it works!)

This it to create what the user sees (What I am envisioning).

The sidebar has:

  - genre dropdown
  - x-axis dropdown
  - y-axis dropdown
  - popularity slider

The main panel has:

  - scatterplot
  - top songs table

```{r build the shiny}
fluidPage(
  titlePanel("Interactive Spotify Track Explorer"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput(
        inputId = "genre",
        label = "Choose a genre:",
        choices = sort(unique(spotify$track_genre)),
        selected = sort(unique(spotify$track_genre))[1]
      ),
      
      selectInput(
        inputId = "x_feature",
        label = "Choose x-axis feature:",
        choices = feature_choices,
        selected = "danceability"
      ),
      
      selectInput(
        inputId = "y_feature",
        label = "Choose y-axis feature:",
        choices = feature_choices,
        selected = "energy"
      ),
      
      sliderInput(
        inputId = "popularity_range",
        label = "Choose popularity range:",
        min = 0,
        max = 100,
        value = c(0, 100)
      )
    ),
    
    mainPanel(
      plotOutput("spotify_plot"),
      br(),
      h3("Top Songs in Selected Genre"),
      DT::DTOutput("top_songs")
    )
  )
)
```

For this part, I had to check some suggestions from Claude to see how it works (cause I am new to this). 

```{r working with the server}
server <- function(input, output, session) {
  
  # This reactive object filters the dataset based on the user's choices.
  # Every time the user changes the genre or popularity range, this dataset updates.
  
  filtered_spotify <- reactive({
    spotify %>%
      filter(
        track_genre == input$genre,
        popularity >= input$popularity_range[1],
        popularity <= input$popularity_range[2]
      )
  })
  
  # This code creates the scatterplot.
  # The x-axis and y-axis change depending on what the user selects.
  
  output$spotify_plot <- renderPlot({
    ggplot(
      filtered_spotify(),
      aes(
        x = .data[[input$x_feature]],
        y = .data[[input$y_feature]],
        color = popularity
      )
    ) +
      geom_point(alpha = 0.6) +
      labs(
        title = paste("Spotify Tracks in the", input$genre, "Genre"),
        x = names(feature_choices)[feature_choices == input$x_feature],
        y = names(feature_choices)[feature_choices == input$y_feature],
        color = "Popularity"
      ) +
      theme_minimal()
  })
  
  # This code creates a table of the most popular songs in the selected genre.
  # I show only the most useful columns so the table stays readable.
  
  output$top_songs <- DT::renderDT({
    filtered_spotify() %>%
      arrange(desc(popularity)) %>%
      select(
        track_name,
        artists,
        album_name,
        popularity,
        danceability,
        energy,
        valence,
        tempo
      ) %>%
      head(10)
  })
}
```

### Run the app 

Let's try to run it!

```{r run it}
shinyApp(ui = fluidPage(
  titlePanel("Interactive Spotify Track Explorer"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput(
        inputId = "genre",
        label = "Choose a genre:",
        choices = sort(unique(spotify$track_genre)),
        selected = sort(unique(spotify$track_genre))[1]
      ),
      
      selectInput(
        inputId = "x_feature",
        label = "Choose x-axis feature:",
        choices = feature_choices,
        selected = "danceability"
      ),
      
      selectInput(
        inputId = "y_feature",
        label = "Choose y-axis feature:",
        choices = feature_choices,
        selected = "energy"
      ),
      
      sliderInput(
        inputId = "popularity_range",
        label = "Choose popularity range:",
        min = 0,
        max = 100,
        value = c(0, 100)
      )
    ),
    
    mainPanel(
      plotOutput("spotify_plot"),
      br(),
      h3("Top Songs in Selected Genre"),
      DT::DTOutput("top_songs")
    )
  )
), server = server)
```

### I couldn't really figure out how to have it show though when knitted (:/). As it is right now, this app works only directly inside the rmd file.  

### Overview 


Ok this was very fun to do! 

This project helped me practice creating an interactive data product instead of only making a static graph. The app allows users to explore Spotify tracks by genre and compare different audio features. For example, users can compare danceability and energy, or look at whether more popular songs tend to have higher valence or tempo.

One thing I learned is that a Shiny app needs both data analysis and design thinking. I had to decide which variables were useful, which controls would be easy for users to understand, and how much information to show at one time. I ended up keping the app simple by using a few dropdowns, one plot, and one table.

I think as I learn more about this with more time, I would like to add more features such as comparing two genres, adding a search box for artists, or creating a recommendation tool based on the user's preferred audio features. And also try to figure out how to have the app run outside the rmd. 