---
title: "Portfolio 10: Creating Figures From Eye-Tracking Data for Poster Presentation"
---

> Goal
For this portfolio piece, I wanted to practice creating figures that I want to use for my first year poster presentation. In the previous portfolio piece, I focused on understanding the eye-tracking dataset, creating a cleaner trial-level and fixation-level dataset, and looking at what the data is looking like. For this piece, I wanted to build on that understanding to make more polished visualizations that summarize the main patterns in the data. 

Here are the figures I want to show in my poster:

1. **Response time cost by block**: Were participants slower when a singleton distractor was present?
2. **Oculomotor capture by block**: How often did the first fixation land on the singleton distractor?
3. **Dwell time by block**: When participants looked at the singleton, how long did they stay on it?

These figures are useful because they show change over time. The broader research question is whether participants learned to avoid or disengage from distracting singleton colors across blocks.

## Load Packages

```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(scales)
```

## Read the Data

```{r read-data}
fix_raw <- read_tsv("data/fix_report.txt", na = c(".", "NA"), show_col_types = FALSE) %>%
  clean_names() %>%
  rename_with(~ str_remove(.x, "^t_"))

glimpse(fix_raw)
```

## Minimal Cleaning and Relabeling

The original dataset is fixation-level, meaning a single trial can appear on multiple rows if the participant made multiple fixations. I convert key variables to numeric, remove practice trials, create readable singleton-condition labels, and keep only the experimental phase.

```{r clean-data}
fix_clean <- fix_raw %>%
  mutate(
    subjnum = as.numeric(subjnum),
    block = as.numeric(block),
    block_norm = block - 1,
    trial = as.numeric(trial),
    acc = as.numeric(acc),
    rt = as.numeric(rt),
    dwell = as.numeric(current_fix_duration),
    targ_loc = as.numeric(targ_loc),
    sing_loc = as.numeric(sing_loc),
    current_fix_x = as.numeric(current_fix_x),
    current_fix_y = as.numeric(current_fix_y),
    current_fix_index = as.numeric(current_fix_index),
    current_fix_start = as.numeric(current_fix_start),
    sing_prob = case_when(
      sing_pres == "A" ~ "Absent",
      sing_prob == "high" ~ "High",
      sing_prob == "low" ~ "Low",
      TRUE ~ as.character(sing_prob)
    ),
    sing_loc = if_else(sing_pres == "A", 0, sing_loc),
    keep_rt = rt > 200 & rt <= 2500
  ) %>%
  filter(
    phase == "experiment",
    block_norm >= 1,
    block_norm <= 14,
    !is.na(subjnum),
    !is.na(trial),
    !is.na(sing_prob)
  )
```

## Helper Function for Plot Summaries

For the poster-style figures, I want one mean value per condition and block. This helper function calculates the mean and standard error.

```{r helper-function}
summary_se <- function(data, value_col, group_cols) {
  data %>%
    group_by(across(all_of(group_cols))) %>%
    summarise(
      mean = mean({{ value_col }}, na.rm = TRUE),
      sd = sd({{ value_col }}, na.rm = TRUE),
      n = n(),
      se = sd / sqrt(n),
      lower = mean - se,
      upper = mean + se,
      .groups = "drop"
    )
}
```

## Figure Theme

I used one consistent theme across the figures so that they look like they belong together.

```{r figure-theme}
poster_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

condition_colors <- c(
  "High" = "#1F9EB7",
  "Low" = "#F4B400",
  "Absent" = "#8ECAE6"
)
```

# Figure 1: Response Time Cost by Block

## Why This Figure?

Response time cost compares singleton-present trials to singleton-absent trials. If participants are slower when the singleton is present, that suggests the distractor made the task harder. I calculated this separately for high-frequency and low-frequency singleton trials.

For each participant and block, I calculated:

- **High RT cost** = High singleton RT - Absent RT
- **Low RT cost** = Low singleton RT - Absent RT

```{r rt-cost-data}
trial_data <- fix_clean %>%
  distinct(subjnum, trial, .keep_all = TRUE) %>%
  filter(
    keep_rt,
    acc == 1,
    !is.na(rt),
    !is.na(sing_prob)
  )

rt_by_subject_block <- trial_data %>%
  group_by(subjnum, block_norm, sing_prob) %>%
  summarise(mean_rt = mean(rt, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sing_prob, values_from = mean_rt) %>%
  filter(!is.na(Absent)) %>%
  mutate(
    High = High - Absent,
    Low = Low - Absent
  ) %>%
  select(subjnum, block_norm, High, Low) %>%
  pivot_longer(
    cols = c(High, Low),
    names_to = "singleton_frequency",
    values_to = "rt_cost"
  ) %>%
  filter(!is.na(rt_cost))

rt_cost_summary <- summary_se(
  data = rt_by_subject_block,
  value_col = rt_cost,
  group_cols = c("block_norm", "singleton_frequency")
)

rt_cost_summary
```

```{r rt-cost-plot, fig.width=9, fig.height=5.5}
ggplot(rt_cost_summary, aes(
  x = block_norm,
  y = mean,
  color = singleton_frequency,
  fill = singleton_frequency
)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  scale_x_continuous(breaks = 1:14) +
  labs(
    title = "Singleton Presence Cost Across Blocks",
    subtitle = "Response time cost = singleton-present RT minus singleton-absent RT",
    x = "Block",
    y = "RT cost (ms)"
  ) +
  poster_theme
```

## Interpretation

This figure shows whether singleton-present trials were slower than singleton-absent trials. Values above zero mean participants were slower when the singleton was present. If the lines decrease across blocks, that would suggest participants became less distracted over time.

# Classifying Fixations for Eye-Tracking Figures

The next two figures require knowing whether a fixation landed on the target, singleton, or a nonsingleton item. The raw data contain fixation coordinates, but not a simple label for the item being fixated. I approximate the item by comparing each fixation coordinate to the eight possible item locations.

```{r classify-fixations}
screen_width <- 1920
screen_height <- 1080
center_x <- screen_width / 2
center_y <- screen_height / 2
eccentricity <- 400
fix_zone <- 90
set_size <- 8

item_locations <- tibble(
  item_loc = 1:set_size,
  angle = (item_loc - 1) * 2 * pi / set_size,
  item_x = center_x + eccentricity * sin(angle),
  item_y = center_y - eccentricity * cos(angle)
)

fix_items <- fix_clean %>%
  filter(
    keep_rt,
    acc == 1,
    !is.na(current_fix_x),
    !is.na(current_fix_y),
    !is.na(dwell)
  ) %>%
  mutate(row_id = row_number()) %>%
  crossing(item_locations) %>%
  mutate(distance = sqrt((current_fix_x - item_x)^2 + (current_fix_y - item_y)^2)) %>%
  group_by(row_id) %>%
  slice_min(distance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    curr_loc = if_else(distance <= fix_zone, item_loc, NA_integer_),
    curr_item = case_when(
      is.na(curr_loc) ~ NA_character_,
      curr_loc == targ_loc ~ "Target",
      curr_loc == sing_loc ~ "Singleton",
      TRUE ~ "Nonsingleton"
    )
  )
```

# Figure 2: Oculomotor Capture by Block

## Why This Figure?

Oculomotor capture asks whether the eyes were pulled to the singleton distractor. For this figure, I looked at the first fixation that landed on a display item on each trial. Then I calculated the percentage of trials where that first fixation landed on the singleton.

```{r oculomotor-capture-data}
first_display_fix <- fix_items %>%
  filter(
    sing_prob %in% c("High", "Low"),
    !is.na(curr_item)
  ) %>%
  group_by(subjnum, trial) %>%
  arrange(current_fix_start, current_fix_index, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(first_on_singleton = curr_item == "Singleton")

oculomotor_by_subject_block <- first_display_fix %>%
  group_by(subjnum, block_norm, sing_prob) %>%
  summarise(
    oculomotor_capture = mean(first_on_singleton, na.rm = TRUE) * 100,
    .groups = "drop"
  )

oculomotor_summary <- summary_se(
  data = oculomotor_by_subject_block,
  value_col = oculomotor_capture,
  group_cols = c("block_norm", "sing_prob")
)

oculomotor_summary
```

```{r oculomotor-capture-plot, fig.width=9, fig.height=5.5}
ggplot(oculomotor_summary, aes(
  x = block_norm,
  y = mean,
  color = sing_prob,
  fill = sing_prob
)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  scale_x_continuous(breaks = 1:14) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Oculomotor Capture Across Blocks",
    subtitle = "Percent of first display fixations landing on the singleton",
    x = "Block",
    y = "First fixations on singleton"
  ) +
  poster_theme
```

## Interpretation

This plot shows how often participants first looked at the singleton distractor. A decrease across blocks would suggest that participants became better at avoiding the singleton. This is related to the idea of oculomotor suppression.

# Figure 3: First Singleton Fixation Dwell Time by Block

## Why This Figure?

The previous figure asks whether the singleton captured the eyes. This figure asks what happened after the singleton was fixated. Dwell time measures how long participants stayed on the singleton once they looked at it. Shorter dwell times may suggest faster disengagement.

```{r dwell-block-data}
first_singleton_fix <- fix_items %>%
  filter(
    sing_prob %in% c("High", "Low"),
    curr_item == "Singleton",
    !is.na(dwell)
  ) %>%
  group_by(subjnum, trial) %>%
  arrange(current_fix_start, current_fix_index, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup()

dwell_by_subject_block <- first_singleton_fix %>%
  group_by(subjnum, block_norm, sing_prob) %>%
  summarise(
    dwell_time = mean(dwell, na.rm = TRUE),
    .groups = "drop"
  )

dwell_summary <- summary_se(
  data = dwell_by_subject_block,
  value_col = dwell_time,
  group_cols = c("block_norm", "sing_prob")
)

dwell_summary
```

```{r dwell-block-plot, fig.width=9, fig.height=5.5}
ggplot(dwell_summary, aes(
  x = block_norm,
  y = mean,
  color = sing_prob,
  fill = sing_prob
)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values = condition_colors) +
  scale_x_continuous(breaks = 1:14) +
  labs(
    title = "First Singleton Fixation Dwell Time Across Blocks",
    subtitle = "Mean duration of the first fixation on the singleton",
    x = "Block",
    y = "Dwell time (ms)"
  ) +
  poster_theme
```
